

## Introduction:
This script takes in input in the form of tab-delimited text files and generates a box-plot displaying the number of differentially expressed genes when using different subsets of samples. The output of this script is a summary plot.  

Load required libraries:

```{r, messages = FALSE}
require(ggplot2)
require(hash)
require(ggpubr)
```

Read in data:

```{r, message = FALSE}
## Read in data for 3 vs 3 comparison:
input_3_vs_3 <- read.table(file = "./input/3_v_3/combined_counts_per_row.txt",
                           col.names = c("test_name",
                                         "de_gene_counts"))

input_3_vs_3$status <- "3_vs_3"
summary(input_3_vs_3$de_gene_counts)

## Read in data for 4 vs 4 comparison:
input_4_vs_4 <- read.table(file = "./input/4_vs_4/combined_counts_per_row.txt",
                           col.names = c("test_name",
                                         "de_gene_counts"))
input_4_vs_4$status <- "4_vs_4"
summary(input_4_vs_4$de_gene_counts)

## Read in data for 5 vs 5 comparison:
input_5_vs_5 <- read.table(file = "./input/5_vs_5/combined_counts_per_row.txt",
                           col.names = c("test_name",
                                         "de_gene_counts"))
input_5_vs_5$status <- "5_vs_5"
summary(input_5_vs_5$de_gene_counts)

## Read in data for 6 vs 6 comparison:
input_6_vs_6 <- read.table(file = "./input/6_vs_6/combined_counts_per_row.txt",
                           col.names = c("test_name",
                                         "de_gene_counts"))
input_6_vs_6$status <- "6_vs_6"
summary(input_6_vs_6$de_gene_counts)
```

For plotting, combine input files:

```{r, message = FALSE}
## Combine:
combined_df <- rbind(input_3_vs_3,
                     input_4_vs_4,
                     input_5_vs_5,
                     input_6_vs_6)

hist_plot <- ggplot(data = combined_df,
                    aes(x = de_gene_counts)) +
        xlab("Number of DEGs") +
        geom_histogram(binwidth = 10) +
        theme_bw()

combined_hist <- hist_plot + facet_wrap(~status)

## Plot
combined_boxplot <- ggplot(combined_df, aes(x = status,
                                            y = de_gene_counts)) +
        xlab("Comparative tests") +
        ylab("Number of DEGs") +
        geom_boxplot() +
        theme_bw()

## Generate a combined plot:
ggarrange(combined_boxplot,
          combined_hist,
          labels = c("A", "B"),
          ncol = 1,
          nrow = 2,
          align = "hv")

## Save to output:
ggsave(file = "combined_comparative_plot.png",
       height = 10,
       width = 10)
```

Perform one sample t-test to explore if the average number of DEGs detected post-diapause using 3 x 3 comparison was significantly different from the number of DEGs detected for infected queens during diapause. 

```{r, message = FALSE}
## Run one sample t test
diapause_degs <- 10
t.test(input_3_vs_3$de_gene_counts,
       mu = diapause_degs)
```
Run lintr to check style:

```{r, message = FALSE}
## Run lintr:
lintr::lint(file = "./plot_de_genes.Rmd")
```
