---
title: "Bombus haemolymph diapause study"
output: plot_go_barchart.html
authors: Joe Colgan (joscolgan)
---

## Introduction:  

This script is for visualisation of the output of gene ontology (GO) analyses using the R package TopGo.
This script creates a barchart of the number of significant genes annotated to a particular GO term, with an associated p-value. This script takes an input of .csv files containing an output table of GO analyses.

```{r, message = FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "lattice", "ggpubr")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        library(lib, character.only = TRUE )
    }
}

## Create output directory:
dir.create("results")
```

## Step One: Load input files
Read in files for the output of Gene Ontology enrichment analysis.

Load results from diapause samples:

```{r, message = FALSE}
## Read in biological processes GO terms:
bp_terms_diapause <- read.table(file = "input_diapause/BP_sig.tsv",
                                header = TRUE)

## Read in molecular function GO terms:
mf_terms_diapause <- read.table(file = "input_diapause/MF_sig.tsv",
                                header = TRUE)

## Read in molecular function GO terms:
cc_terms_diapause <- read.table(file = "input_diapause/CC_sig.tsv",
                                header = TRUE)
```

Read in postdiapause samples:

```{r, message = FALSE}
## Read in biological processes GO terms:
bp_terms_pd <- read.table(file = "results/output_node_20_raw/BP_sig.tsv",
                       header = TRUE)

## Read in molecular function GO terms:
mf_terms_pd <- read.table(file = "results/output_node_20_raw/MF_sig.tsv",
                       header = TRUE)

## Read in molecular function GO terms:
cc_terms_pd <- read.table(file = "results/output_node_20_raw/CC_sig.tsv",
                       header = TRUE)

```

For each time-point, combine gene ontolgy terms for plotting:

```{r, message = FALSE}
## Combine GO terms:
combined_terms_diapause <- rbind(bp_terms_diapause,
                        mf_terms_diapause,
                        cc_terms_diapause)
combined_terms_diapause$category <- factor(combined_terms_diapause$category)

## Combine GO terms:
combined_terms_pd <- rbind(bp_terms_pd,
                        mf_terms_pd,
                        cc_terms_pd)
combined_terms_pd$category <- factor(combined_terms_pd$category)

## Set order of GO categories to plot
levels(combined_terms_diapause$category) <- c("BP", "MF", "CC")
levels(combined_terms_pd$category) <- c("BP", "MF", "CC")

## Log transform p values for plotting:
combined_terms_diapause$weight_ks_adjusted <- -log(combined_terms_diapause$weight_ks_adjusted)
combined_terms_pd$weight_ks_adjusted <- -log(combined_terms_pd$weight_ks_adjusted)

## Reorder terms for plotting:
combined_terms_diapause$Term <- factor(combined_terms_diapause$Term,
                                       levels = combined_terms_diapause$Term[order(combined_terms_diapause$category,                                                                                          combined_terms_diapause$weight_fisher_adjusted)])

combined_terms_pd$Term <- factor(combined_terms_pd$Term,
                                 levels = combined_terms_pd$Term[order(combined_terms_pd$category,                                      combined_terms_pd$weight_fisher_adjusted)])

## For plotting, update GO term names to include total number of annotated terms:
combined_terms_diapause$updated_terms <- paste(combined_terms_diapause$Term,
                                      " ",
                                      "(",
                                      combined_terms_diapause$Annotated,
                                      ")",
                                      sep = "")

combined_terms_pd$updated_terms <- paste(combined_terms_pd$Term,
                                      " ",
                                      "(",
                                      combined_terms_pd$Annotated,
                                      ")",
                                      sep = "")

## Remove underscore:
combined_terms_diapause$updated_terms <- gsub("_", " ",
                                              combined_terms_diapause$updated_terms)
combined_terms_pd$updated_terms <- gsub("_", " ",
                                        combined_terms_pd$updated_terms)

## Generate plot:
combined_terms_diapause_plot <- ggbarplot(combined_terms_diapause,
                                          x = "updated_terms",
                                          y = "weight_ks_adjusted",
                                          position = position_dodge(0.1),
                                          fill = "category",
                                          color = NULL,
                                          palette = "jco",
                                          sort.val = "asc",
                                          sort.by.groups = TRUE,
                                          ylab = "-log10(p)",
                                          xlab = "Gene ontology term",
                                          legend.title = "Gene ontology",
                                          lab.col = "black",
                                          lab.size = 4,
                                          lab.vjust = 0.5,
                                          lab.hjust = 1,
                                          legend = "top",
                                          rotate = TRUE,
                                          ggtheme = theme_minimal())

## Make font bigger and bold:
combined_terms_diapause_plot <- combined_terms_diapause_plot +
        scale_y_continuous(expand = c(0, 0)) +
        theme(axis.text = element_text(size = 15),
              axis.title.x = element_text(size = 15,
                                          face = "bold"),
              axis.title.y = element_text(size = 15,
                                          face = "bold"),
              axis.text.y = element_text(size = 12,
                                         face = "bold"),
              axis.text.x = element_text(size = 12),
              legend.position = "none") +
        expand_limits(y = 30) +
        geom_hline(yintercept = 1.301,
                   linetype = "dashed",
                   colour = "black")

combined_terms_diapause_plot
## Generate plot:
combined_terms_pd_plot <- ggbarplot(combined_terms_pd,
                                          x = "updated_terms",
                                          y = "weight_ks_adjusted",
                                          position = position_dodge(0.1),
                                          fill = "category",
                                          color = NULL,
                                          palette = "jco",
                                          sort.val = "asc",
                                          sort.by.groups = TRUE,
                                          ylab = "-log10(p)",
                                          xlab = "Gene ontology term",
                                          legend.title = "Gene ontology",
                                          lab.col = "black",
                                          lab.size = 4,
                                          lab.vjust = 0.5,
                                          lab.hjust = 1,
                                          legend = "top",
                                          rotate = TRUE,
                                          ggtheme = theme_minimal())

## Make font bigger and bold:
combined_terms_pd_plot <- combined_terms_pd_plot +
        scale_y_continuous(expand = c(0, 0)) +
        theme(axis.text = element_text(size = 15),
              axis.title.x = element_text(size = 15,
                                          face = "bold"),
              axis.title.y = element_text(size = 15,
                                          face = "bold"),
              axis.text.y = element_text(size = 12,
                                         face = "bold"),
              axis.text.x = element_text(size = 12),
              legend.position = "none") +
        expand_limits(y = 30) +
        geom_hline(yintercept = 1.301,
                   linetype = "dashed",
                   colour = "black")

## Print together:
ggarrange(combined_terms_diapause_plot,
          combined_terms_pd_plot,
          ncol = 1,
          nrow = 2,
          align = "hv",
          heights = c(1.0, 2.2),
          labels = c("A", "B"))

## Save picture:
ggsave(file = "results/combined_go_terms_barchart.png",
       height = 10,
       width = 20)
```

Run lintr:

```{r, message = FALSE}
## Run lintr
lintr::lint(file = "./plot_go_barchart.Rmd")
```