---
title: "Sphaerularia-Bombus infection RNA-Seq study"
output: GO_enrichment_visualisation.html
---

## Introduction 
The purpose of this script is to read in a list of significant gene ontology terms enriched with genes differentially
expressed within bumblebee (Bombus terrestris) queens in response to infection by the nematode, Sphaerularia bombi. 

The first step involves loading (and installing if required) R packages:
```{r, message = FALSE}
# Load libraries; install from scratch if needed
libraries <- c("ggplot2", "lattice", "ggpubr")
for (lib in libraries) {
    if (require(package = lib, character.only = TRUE)) {
        print("Successful")
    } else {
        print("Installing")
        source("https://bioconductor.org/biocLite.R")
        biocLite(pkgs = lib)
        library(lib, character.only = TRUE )
    }
}
```

The study consists of two independent datasets:
- Dataset 1: Bumblebee queens collected during diapause (infected (n=3) vs. control (n=3))
- Dataset 2: Bumblebee queens collected post-diapause (infected (n=7) vs. control (n=7))

```{r}
### Plotting of enriched gene ontology terms during diapause:
## Load in the data:
diapause_deseq_BP <- read.table(file="results/output_node_20_diapause/BP_sig.txt", 
                                           header = TRUE)

diapause_deseq_MF <- read.table(file="results/output_node_20_diapause/MF_sig.txt", 
                                           header = TRUE)

diapause_deseq_CC <- read.table(file="results/output_node_20_diapause/CC_sig.txt", 
                                           header = TRUE)

## Combine the individual list of terms:
diapause_deseq_combined <- rbind(diapause_deseq_BP, 
                                 diapause_deseq_MF,
                                 diapause_deseq_CC)

## Filter out terms with BH adjusted p values:
diapause_deseq_combined_sig <- subset(diapause_deseq_combined, weight_ks_adjusted < 0.05)


diapause_deseq_combined_sig$category <- factor(diapause_deseq_combined_sig$category)

# Set order of GO categories to plot
levels(diapause_deseq_combined_sig$category) <- c("BP", "MF", "CC")

# Round p values to plot
diapause_deseq_combined_sig$weight_ks_adjusted <- round(diapause_deseq_combined_sig$weight_ks_adjusted, digits = 3)

# Log p values for plotting 
diapause_deseq_combined_sig$weight_ks_adjusted <- -log(diapause_deseq_combined_sig$weight_ks_adjusted)

# Reorder terms for plotting 
diapause_deseq_combined_sig$Term <- factor(diapause_deseq_combined_sig$Term, levels = diapause_deseq_combined_sig$Term[order(diapause_deseq_combined_sig$category, diapause_deseq_combined_sig$weight_ks_adjusted)])

## For plotting queen, update GO term names to include total number of annotated terms:
diapause_deseq_combined_sig$updated_terms <- paste(diapause_deseq_combined_sig$Term, " ", "(", diapause_deseq_combined_sig$Annotated, ")", sep="")

## Remove underscores:
diapause_deseq_combined_sig$updated_terms <- gsub("_", " ", diapause_deseq_combined_sig$updated_terms)

## Generate a dot plot for all enriched terms within infected diapause queens:
diapause_deseq_combined_plot <- ggdotchart(data = diapause_deseq_combined_sig, 
                                           x = "updated_terms", y = "weight_ks_adjusted",
                                           color = "category",
                                           palette = c("#00AFBB", "#E7B800", "#FC4E07"),
                                           sorting = "descending",
                                           ylab = "-log10(p)",
                                           xlab = "Gene ontology term",
                                           legend.title = "Gene ontology",
                                           add = "segments",
                                           rotate = TRUE,
                                           group = "category",
                                           dot.size = 6,
                                           ggtheme = theme_pubr()
                                           )
                                               
## Update the size of labels:
diapause_deseq_combined_plot <- diapause_deseq_combined_plot +
                                scale_y_continuous(expand = c(0, 0)) +
                                theme(axis.text=element_text(size=10),
                                      axis.title.x = element_text(size=15,face="bold"),
                                      axis.title.y = element_text(size=15,face="bold"),
                                      axis.text.y = element_text(size=15, face="bold"),
                                      axis.text.x = element_text(size=15),
                                      legend.position="none") +
                                expand_limits(y = 10) +
                                geom_hline(yintercept = 1.301, linetype="dashed", colour="black")

##
ggsave("diapause_deseq_combined_go_term_plot.png", 
       height = 7.5,
       width = 12.5)
```

For the second dataset:

```{r}
### Plotting of enriched gene ontology terms post-diapause:
## Load in the data:
postdiapause_deseq_BP <- read.table(file="results/output_node_20_postdiapause/BP_sig.txt", 
                                           header = TRUE)

postdiapause_deseq_MF <- read.table(file="results/output_node_20_postdiapause/MF_sig.txt", 
                                           header = TRUE)

postdiapause_deseq_CC <- read.table(file="results/output_node_20_postdiapause/CC_sig.txt", 
                                           header = TRUE)

## Combine the individual list of terms:
postdiapause_deseq_combined <- rbind(postdiapause_deseq_BP, 
                                     postdiapause_deseq_MF,
                                     postdiapause_deseq_CC)

## Filter out terms with BH adjusted p values:
postdiapause_deseq_combined_sig <- subset(postdiapause_deseq_combined, weight_ks_adjusted < 0.05)


postdiapause_deseq_combined_sig$category <- factor(postdiapause_deseq_combined_sig$category)

# Set order of GO categories to plot
levels(postdiapause_deseq_combined_sig$category) <- c("BP", "MF", "CC")

# Round p values to plot
postdiapause_deseq_combined_sig$weight_ks_adjusted <- round(postdiapause_deseq_combined_sig$weight_ks_adjusted, digits = 3)

# Log p values for plotting 
postdiapause_deseq_combined_sig$weight_ks_adjusted <- -log(postdiapause_deseq_combined_sig$weight_ks_adjusted)

# Reorder terms for plotting 
postdiapause_deseq_combined_sig$Term <- factor(postdiapause_deseq_combined_sig$Term, levels = postdiapause_deseq_combined_sig$Term[order(postdiapause_deseq_combined_sig$category, postdiapause_deseq_combined_sig$weight_ks_adjusted)])

## For plotting queen, update GO term names to include total number of annotated terms:
postdiapause_deseq_combined_sig$updated_terms <- paste(postdiapause_deseq_combined_sig$Term, " ", "(", postdiapause_deseq_combined_sig$Annotated, ")", sep="")

## Remove underscores:
postdiapause_deseq_combined_sig$updated_terms <- gsub("_", " ", postdiapause_deseq_combined_sig$updated_terms)

## Generate a dot plot for all enriched terms within infected diapause queens:
postdiapause_deseq_combined_plot <- ggdotchart(data = postdiapause_deseq_combined_sig, 
                                           x = "updated_terms", y = "weight_ks_adjusted",
                                           color = "category",
                                           palette = c("#00AFBB", "#E7B800", "#FC4E07"),
                                           sorting = "descending",
                                           ylab = "-log10(p)",
                                           xlab = "Gene ontology term",
                                           legend.title = "Gene ontology",
                                           add = "segments",
                                           rotate = TRUE,
                                           group = "category",
                                           dot.size = 6,
                                           ggtheme = theme_pubr()
                                           )
                                               
## Update the size of labels:
postdiapause_deseq_combined_plot <- postdiapause_deseq_combined_plot +
                                scale_y_continuous(expand = c(0, 0)) +
                                theme(axis.text=element_text(size=10),
                                      axis.title.x = element_text(size=15,face="bold"),
                                      axis.title.y = element_text(size=15,face="bold"),
                                      axis.text.y = element_text(size=15, face="bold"),
                                      axis.text.x = element_text(size=15),
                                      legend.position="none") +
                                expand_limits(y = 10) +
                                geom_hline(yintercept = 1.301, linetype="dashed", colour="black")
```

Combine plots:

```{r}
## Combine plot:
ggarrange(diapause_deseq_combined_plot, postdiapause_deseq_combined_plot,
                           align = "hv",
                           nrow = 2, ncol = 1,
                           heights = c(0.5,1.2),
                           common.legend = TRUE,
                           #labels = c("A) Diapause", "B) Postdiapause"),
                           vjust = c(1.4,1.4))


ggsave("diapause_deseq_combined_go_term_plot.png", 
       height = 7.5,
       width = 12.5)
```


